The Form of Uncertainty Affects Selection for Social Learning

### 2023-04-28 ms.tex
final version for publication

\begin{document}
\maketitle
\clearpage

@@

All agents perform individual, trial-and-error learning at each time step in
their lifespan.  Learning is guided by softmax search. Softmax search
guides agents to exploit more frequently the most profitable behaviors when the
agent is more certain it is the optimal one and to explore more frequently when the
agent is unsure. To do softmax searching, agents track average payoffs acquired
from each behavior they have performed, which requires knowing the number of times they have performed each behavior. The probability that the agent will perform a behavior is a function of the agent's beliefs about a behavior's average payoff in that time step and a fixed parameter that influences the amount of exploratory behavior (more details below). The softmax function used here is a biologically plausible generalization of behavior selection under uncertainty that enables agents to often greedily exploit the most lucrative behavior they have observed, but also to sometimes explore alternatives~\citep{Schulz2019,Collins2013,Daw2006,Yechiam2005}.

@@

The model is initialized with $N$ agents. Each agent $i$ tracks 
observed mean payoffs for each behavior, $b$, which is denoted $\ledger$. Each
agent $i$ also counts how many times they have tried each behavior $b$, 
denoted $c_{ib}$. At model initialization, $\ledger = 0.0$ and $c_{ib} = 0$ for
all $i$ and $b$.% (Figure~\ref{fig:schematic}A).
One of the behaviors is chosen at random to yield expected payoff $\pihigh$, with the other $B-1$ behaviors yielding
expected payoffs $\pilow$. Agents are independently randomly initialized with $s_i
\in 0,1$, so that $\sim$50\% of the initial population learn socially.

@@

Of the four types of uncertainty studied in this paper, payoff ambiguity had the most
complex effect on the evolution of social learning. Recall that we operationalized
payoff ambiguity with $\pilow$, the expected payoff for non-optimal behaviors
(equivalently the probability that those behaviors yielded a payoff of 1). This was
relative to $\pihigh = 0.9$, the expected payoff for the optimal behavior. When
payoff ambiguity was low, non-optimal behaviors rarely paid off, so that exploration
yielded reliable information and selection on learning strategies was strong. When
payoff ambiguity is very high ($\pilow = 0.8$), two things happen. First, agents are
more likely to err by ascribing high value to non-optimal behaviors, since it is
more difficult for them to discern the difference between optimal and sub optimal
behaviors. Second, and perhaps more importantly, natural selection on strategies
that \emph{do} reliably select the optimal behavior is relatively weak. This means
selection is weaker \emph{in favor} of social learning under low values of $u$ and
weaker \emph{against} social learning under high values of $u$ (Figure~\ref{fig:mainResults}, bottom row). 
This effect was especially pronounced under other types of uncertainty, i.e., larger selection-set sizes and shorter effective lifespans---particularly when $L=1$ where no individual learning occurred.

@@

To begin to confirm our interpretation of our analysis above, we measured
strength of selection via the ensemble average of the number of
time steps to fixation over all simulation trials for a given parameter setting. More steps to fixation here indicates weaker selection. Strength of selection is the extent to which the
social-learning trait provided agents with a consistent benefit. Selection
strength is minimal when social and asocial learning
are equally beneficial. 
% Even under consistent but weak selection pressure, drift can cause a population to take longer to reach fixation \citep{plutynski2007drift}. 
We denote the ensemble average over simultaion trials of the number of generations to fixation, denoted $\langle G \rangle$, where fixation means that the number of social learners, $\sum_i s_i$, is equal to either zero or $N$. %Indeed, social learning evolution and drift are reflected in the relative benefit of social learning to asocial learning.

Figure~\ref{fig:steps} shows the average number of generations to fixation across all uncertainty parameters. Drift was greatest overall when payoff ambiguity was greatest, i.e., when $\pilow = 0.8$. When payoffs are highly ambiguous there is relatively little advantage to the optimal learning strategy, and therefore relatively weak selective pressure.

@@

Figure~\ref{fig:payoffs} shows the average net payoffs normalized by lifespan
for our main simulation compared with the two reference homogeneous populations
initialized either as all-social or all-asocial learners. Often the simulated payoffs follow the payoffs of the better-performing homogeneous group. However, there are several important deviations worth explaining. First, populations sometimes evolved to be asocial learners, even
though all-social learner populations outperformed all-individual-learner populations when agents were
long-lived ($L=4$ and $L=8$) and in simple environments ($B=2$; e.g., Figure~\ref{fig:payoffs}A).
Second, we also observed the opposite dynamic: agents from our main simulations who evolved to be social learners sometimes outperformed agents in the homogeneous social learner populations. This is most pronounced when $L=1$, and even more
so when $\pilow=0.45$ and $B=10$ (Figure~\ref{fig:payoffs}C).
In fact, social learning evolved even in parts of the parameter space where the expected homogeneous social-learning payoff was lower than the expected homogeneous
asocial-learning payoff (Figure~\ref{fig:payoffs}B, e.g., $L=1$).

@@

When payoff ambiguity and selection-set size were small ($\pilow=0.1$ and $B=2$), and lifespan
was long ($L \geq 4$), asocial learning fixated because periods of extended
environmental instability (see environmental changes after generation 50 in  Figure~\ref{fig:payoffTimeseries}A) caused social learners to more severely underperform compared
to asocial learners, even though expected
homogeneous social payoffs were greater than homogeneous asocial payoffs. Runs of environmental change strongly affected the geometric mean fitness of social learners who were weighing
outdated information too heavily. In other words, environmental change lowers social learners'
mean payoffs, but social-learner payoffs build back up as the information quality in the
population improves (blue dots, Figure~\ref{fig:payoffTimeseries}A). Even though a homogenous
population of social learners would outperform individual learners in this case, unpredictable
environmental stochasticity undermines the reliability of social learning. When lifespan was
minimal ($L=1$) and environmental variability was moderate ($u\approx0.5$), the opposite,
unexpected outcome could occur where social learning fixates despite the fact that a homogenous
population of asocial learners would outperform a homogenous population of social learners
(Figure~\ref{fig:payoffTimeseries}B). In this case, it takes relatively few consecutive generations with environmental stability to attract the population to all become social learners since social learners greatly outperform asocial learners during longer periods of environmental stability.
\begin{figure}
  \caption{Average number of generations (y-axes) to fixation. 
    Note the y-axis scale varies between plots, indicating different overall 
    time to fixation for different $(\pilow, B)$ combinations.} 
  \label{fig:steps}
\centering
    \includegraphics[width=\textwidth]{Figures/stepResultsPlots.pdf}
\end{figure}

@@

By carefully identifying and operationalizing different forms of uncertainty common in social-learning models, we developed a more thorough understanding of how uncertainty can favor social learning and how different forms of uncertainty interact to favor its evolution. By explicitly modeling a key individual-level learning mechanism shared across taxa, namely the ability to adjust behavior to uncertainty and new information, we found that social learning could evolve even if it provides outdated information. %Social learning is a critical component of problem solving among social creatures. Our work identified and systematized some important forms of uncertainty and catalogued their effects on the evolution of social learning.
Our work, then, both advances the theory of social learning and could eventually be of practical use in identifying specific mechanisms underlying adaptations to uncertain environments.

\paragraph{Supplementary material.} Supplemental analyses confirming model robustness to free parameters is available at \url{https://osf.io/tpsy9}. Model and analysis code is
publicly available on GitHub (\url{https://github.com/mt-digital/UncMod}), and the data used to make the figures presented here is available via OSF
(\url{https://osf.io/t8exa/}). A guide to running the analyses is available on the GitHub repository.

@@

\paragraph{Author contributions.} JHJ, CM, PES, and MAT conceived of the project. MAT developed the model
and analyses, with contributions from CM, PES, and JHJ. MAT implemented and analyzed the model. MAT wrote and edited the initial manuscript draft, which was extensively expanded and edited by CM, PES, and JHJ.

@@

@@

\vspace{-3em}
\begin{figure}
  \centering
   % \addtocounter{figure}{-1}
  \caption{Number of prospective teachers sensitivity analysis for $N_T=2,20$. Recall
  $N_T=5$ was used to generate main text results.}
  \label{fig:nteachersSensitivity}
  \vspace{2em}
  \begin{subfigure}{\textwidth}
	\caption{Number of prospective teachers, $N_T = 2$}
	\includegraphics[width=\textwidth]{Figures/supplement/nteachers=2/mainResultsPlots.pdf}
  \end{subfigure}
\end{figure}
% \newpage
% \begin{figure}
%   \ContinuedFloat
%   \begin{subfigure}{\textwidth}
% 	\caption{$N_T = 10$}
% 	\includegraphics[width=\textwidth]{Figures/supplement/nteachers=10/mainResultsPlots.pdf}
%   \end{subfigure}
% \end{figure}
\newpage
\begin{figure}
  \ContinuedFloat
  \begin{subfigure}{\textwidth}
	\caption{Number of prospective teachers, $N_T = 20$}
	\includegraphics[width=\textwidth]{Figures/supplement/nteachers=20/mainResultsPlots.pdf}
  \end{subfigure}
\end{figure}

>>>>

\begin{document}
\maketitle
\noindent
\textbf{Social media summary:} You won’t believe this one weird adaptive trick social learners use{\ldots}asocial
learners hate it! \\
\textbf{Keywords:} Uncertainty, social learning, cultural evolution, agent-based modeling \\
\clearpage

@@

All agents perform individual, trial-and-error learning at each time step in
their lifespan.  Learning is guided by softmax search. Softmax search
guides agents to exploit more frequently the most profitable behaviors when the
agent is more certain it is the optimal one and to explore more frequently when the
agent is unsure. To do softmax searching, agents track average payoffs acquired
from each behavior they have performed, which requires knowing the number of times they have performed each
behavior. The probability that an agent will perform a behavior is a function of the agent's beliefs about a
behavior's average payoff in that time step and a fixed parameter that influences the amount of exploratory
behavior (more details below). The softmax function used here is a biologically plausible generalization of
behavior selection under uncertainty that enables agents to often greedily exploit the most lucrative
behavior they have observed, but also to sometimes explore
alternatives~\citep{Yechiam2005,Daw2006,Collins2013,Schulz2019}.

@@

The model is initialized with $N$ agents. Each agent $i$ tracks 
observed mean payoffs for each behavior, $b$, which is denoted $\ledger$. Each
agent $i$ also counts how many times they have tried each behavior $b$, 
denoted $c_{ib}$. At model initialization, $\ledger = 0.0$ and $c_{ib} = 0$ for
all $i$ and $b$. % (Figure~\ref{fig:schematic}A).
One of the behaviors is chosen at random to yield expected payoff $\pihigh$, with the other $B-1$ behaviors yielding
expected payoffs $\pilow$. Agents are independently randomly initialized with $s_i
\in 0,1$, so that $\sim$50\% of the initial population learn socially.

@@

Of the four types of uncertainty studied in this paper, payoff ambiguity had the most
complex effect on the evolution of social learning. Recall that we operationalized
payoff ambiguity with $\pilow$, the expected payoff for non-optimal behaviors
(equivalently the probability that those behaviors yielded a payoff of 1). This was
relative to $\pihigh = 0.9$, the expected payoff for the optimal behavior. When
payoff ambiguity was low, non-optimal behaviors rarely paid off, so that exploration
yielded reliable information and selection on learning strategies was strong. When
payoff ambiguity is very high ($\pilow = 0.8$), two things happen. First, agents are
more likely to err by ascribing high value to non-optimal behaviors, since it is
more difficult for them to discern the difference between optimal and suboptimal
behaviors. Second, and perhaps more importantly, natural selection on strategies
that \emph{do} reliably select the optimal behavior is relatively weak. This means
selection is weaker \emph{in favor} of social learning under low values of $u$ and
weaker \emph{against} social learning under high values of $u$ (Figure~\ref{fig:mainResults}, bottom row). 
This effect was especially pronounced under other types of uncertainty, i.e., larger selection-set sizes and shorter effective lifespans---particularly when $L=1$ where no individual learning occurred.

@@

To begin to confirm our interpretation of our analysis above, we measured
strength of selection via the ensemble average of the number of
time steps to fixation over all simulation trials for a given parameter setting. More steps to fixation here indicates weaker selection. Strength of selection is the extent to which the
social-learning trait provided agents with a consistent benefit. Selection
strength is minimal when social and asocial learning
are equally beneficial. 
% Even under consistent but weak selection pressure, drift can cause a population to take longer to reach fixation \citep{plutynski2007drift}. 
We denote the ensemble average over simultaion trials of the number of generations to fixation as $\langle G \rangle$, where fixation means that the number of social learners, $\sum_i s_i$, is equal to either zero or $N$. %Indeed, social learning evolution and drift are reflected in the relative benefit of social learning to asocial learning.

Figure~\ref{fig:steps} shows the average number of generations to fixation across all uncertainty
parameters. Strength of selection was weakest overall when payoff ambiguity was greatest, i.e., when $\pilow = 0.8$. When payoffs are highly ambiguous there is relatively little advantage to the optimal learning strategy, and therefore relatively weak selective pressure.

@@

Figure~\ref{fig:payoffs} shows the average net payoffs normalized by lifespan
from a selection of our main simulations compared with the two reference homogeneous populations
initialized either as all-social or all-asocial learners. Often the simulated payoffs follow the payoffs of the better-performing homogeneous group. However, there are several important deviations worth explaining. First, populations sometimes evolved to be asocial learners, even
though all-social learner populations outperformed all-individual-learner populations when agents were
long-lived ($L=4$ and $L=8$) and in simple environments ($B=2$; e.g., Figure~\ref{fig:payoffs}A).
Second, we also observed the opposite dynamic: agents from our main simulations who evolved to be social learners sometimes outperformed agents in the homogeneous social learner populations. This is most pronounced when $L=1$, and even more
so when $\pilow=0.45$ and $B=10$ (Figure~\ref{fig:payoffs}C).
In fact, social learning evolved even in parts of the parameter space where the expected homogeneous social-learning payoff was lower than the expected homogeneous
asocial-learning payoff (Figure~\ref{fig:payoffs}B, e.g., $L=1$).

@@

When payoff ambiguity and selection-set size were small ($\pilow=0.1$ and $B=2$), and lifespan
was long ($L \geq 4$), asocial learning fixated because periods of extended
environmental instability (see environmental changes after generation 50 in  Figure~\ref{fig:payoffTimeseries}A) caused social learners to more severely underperform compared
to asocial learners, even though expected
homogeneous social payoffs were greater than homogeneous asocial payoffs. Runs of environmental change strongly affected the geometric mean fitness of social learners who were weighing
outdated information too heavily. In other words, environmental change lowers social learners'
mean payoffs, but social-learner payoffs build back up as the information quality in the
population improves (blue dots, Figure~\ref{fig:payoffTimeseries}A). Even though a homogenous
population of social learners would outperform asocial learners in this case, unpredictable
environmental stochasticity undermines the reliability of social learning. When lifespan was
minimal ($L=1$) and environmental variability was moderate ($u\approx0.5$), the opposite,
unexpected outcome could occur where social learning fixates despite the fact that a homogenous
population of asocial learners would outperform a homogenous population of social learners
(Figure~\ref{fig:payoffTimeseries}B). In this case, it takes relatively few consecutive generations with environmental stability to attract the population to all become social learners since social learners greatly outperform asocial learners during longer periods of environmental stability.
\begin{figure}
  \caption{Average number of generations (y-axes) to fixation. 
    Note the y-axis scale varies between plots, indicating different overall 
    time to fixation for different $(\pilow, B)$ combinations.} 
  \label{fig:steps}
\centering
    \includegraphics[width=\textwidth]{Figures/stepResultsPlots.pdf}
\end{figure}

@@

By carefully identifying and operationalizing different forms of uncertainty common in social-learning
models, we developed a more thorough understanding of how uncertainty can favor social learning and how
different forms of uncertainty interact to favor its evolution. By explicitly modeling a key
individual-level learning mechanism shared across taxa, namely the ability to adjust behavior to uncertainty
and new information, we found that social learning could evolve even if it regularly provides outdated information. %Social learning is a critical component of problem solving among social creatures. Our work identified and systematized some important forms of uncertainty and catalogued their effects on the evolution of social learning.
Our work, then, both advances the theory of social learning and could eventually be of practical use in identifying specific mechanisms underlying adaptations to uncertain environments.

\paragraph{Supplementary material.} Supplemental analyses confirming model robustness to free parameters is available at \url{https://osf.io/tpsy9}.

@@

\paragraph{Author contributions.} JHJ, CM, PES, and MAT conceived of the project. MAT developed the model
and analyses, with contributions from CM, PES, and JHJ. MAT implemented and analyzed the model. MAT wrote and edited the initial manuscript draft, which was expanded and edited by CM, PES, and JHJ.

@@

\paragraph{Research transparency and reproducibility.}
Model and analysis code is publicly available on GitHub (\url{https://github.com/mt-digital/UncMod}).  
A guide to running the analyses is available on the GitHub repository.
The code version used to generate the results presented here has been archived
with a persistent identifier through Zenodo: \url{https://doi.org/10.5281/zenodo.7873161}.

\paragraph{Data availability statement.} 
The data used to make the figures presented here is available via OSF
(\url{https://osf.io/t8exa/}).

@@

\vspace{-3em}
\begin{figure}
  \centering
   % \addtocounter{figure}{-1}
  \caption{Number of prospective teachers sensitivity analysis for $N_T=2,20$. Recall
  $N_T=5$ was used to generate main text results.}
  \label{fig:nteachersSensitivity}
  \vspace{2em}
  \begin{subfigure}{\textwidth}
	\caption{Number of prospective teachers, $N_T = 2$}
	\includegraphics[width=\textwidth]{Figures/supplement/nteachers=2/mainResultsPlots.pdf}
  \end{subfigure}
\end{figure}
% \newpage
% \begin{figure}
%   \ContinuedFloat
%   \begin{subfigure}{\textwidth}
% 	\caption{$N_T = 10$}
% 	\includegraphics[width=\textwidth]{Figures/nteachers=10/mainResultsPlots.pdf}
%   \end{subfigure}
% \end{figure}
\newpage
\begin{figure}
  \ContinuedFloat
  \begin{subfigure}{\textwidth}
	\caption{Number of prospective teachers, $N_T = 20$}
	\includegraphics[width=\textwidth]{Figures/supplement/nteachers=20/mainResultsPlots.pdf}
  \end{subfigure}
\end{figure}
